---
title: "Stats: EDA Basics"
# yaml-author
date: 2020-05-07
output: github_document
time: 20
reading: 45
---

# Stats: EDA Basics

*Purpose*: *Exploratory Data Analysis* (EDA) is a __crucial__ skill for a
practicing data scientist. Unfortunately, much like human-centered design EDA is
hard to teach. This is because EDA is **not** a strict procedure, so much as it
is a **mindset**. Also, much like human-centered design, EDA is an *iterative,
nonlinear process*. There are two key principles to keep in mind when doing EDA:

- 1. Curiosity: Generate lots of ideas and hypotheses about your data.
- 2. Skepticism: Remain unconvinced of those ideas, unless you can find credible
  patterns to support them.

Since EDA is both *crucial* and *difficult*, we will practice doing EDA *a lot*
in this course!

*Reading*: [Exploratory Data Analysis](https://rstudio.cloud/learn/primers/3.1)
*Topics*: (All topics)
*Reading Time*: ~45 minutes

*Note*: This exercise will consist of interpreting pre-made graphs. You can run
the whole notebook to generate all the figures at once. Just make sure to do
all the exercises and write your observations!

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

### __q0__ Remember from `e02-data-basics` there were *simple checks* we're supposed
to do? Do those simple checks on the diamonds dataset below.

```{r q0-task}
glimpse(diamonds)
summary(diamonds)
```

I'm going to walk you through a train of thought I had when studying the
diamonds dataset.

There are four standard "C's" of
[judging](https://en.wikipedia.org/wiki/Diamond_(gemstone)) a diamond.[1] These
are `carat, cut, color` and `clarity`, all of which are in the `diamonds`
dataset.

## Hypothesis 1

**Here's a hypothesis**: `Ideal` is the "best" value of `cut` for a diamond.
Since an `Ideal` cut seems more labor-intensive, I hypothesize that `Ideal` cut
diamonds are less numerous than other cuts.

### __q1__ Run the chunk below, and study the plot. Was hypothesis 1 correct? Why
or why not?

```{r q1-task}
diamonds %>%
  ggplot(aes(cut)) +
  geom_bar()
```

**Observations**:
- Actually, Ideal cut diamonds are the most common cut of diamond in this dataset, although they do not outnumber the rest of the diamonds combined.

## Hypothesis 2

**Another hypothesis**: The `Ideal` cut diamonds should be the most pricey.

### __q2.1__ Study the following graph; does it support, contradict, or not relate to hypothesis 2?

*Hint*: Is this an effective graph? Why or why not?

```{r q2.1-task}
diamonds %>%
  ggplot(aes(cut, price)) +
  geom_point()
```

**Observations**:
- It is very difficult to determine from this graph whether or not ideal cut diamonds are more expensive than the rest of the diamond cuts. Based on the graph, it looks like ideal cut diamonds are essentially the same price, but the graph may be misleading.

The following is a set of *boxplots*; the middle bar denotes the median, the
boxes denote the *quartiles* (upper and lower "quarters" of the data), and the
lines and dots denote large values and outliers.

### __q2.2__ Study the following graph; does it support or contradict hypothesis 2?

```{r q2.2-task}
diamonds %>%
  ggplot(aes(cut, price)) +
  geom_boxplot()
```

**Observations**:
- The graph contradicts hypothesis 2 as the median and upper quartile price for ideal cut diamonds is clearly lower than both premium and very good diamonds in this dataset. However, there are  more outliers in the ideal dataset than very good and premium diamonds, which may indicate that ideal cut diamonds are not necessarily always less expensive.

## Unraveling Hypothesis 2

Upon making the graph in __q2.2__, I was very surprised. So I did some reading
on diamond cuts. It turns out that some gemcutters [sacrifice cut for
carat](https://en.wikipedia.org/wiki/Diamond_(gemstone)#Cut). Could this effect
explain the surprising pattern above?

### __q3__ Study the following graph; does it support a "carat over cut" hypothesis? How might this relate to price?

*Hint*: The article linked above will help you answer these questions!

```{r q2.3-task}
diamonds %>%
  ggplot(aes(cut, carat)) +
  geom_boxplot()
```

**Observations**:
- According to the wikipedia article, "The price per carat does not increase linearly with increasing size. Instead, there are sharp jumps around milestone carat weights, as demand is much higher for diamonds weighing just more than a milestone than for those weighing just less. As an example, a 0.99-carat diamond may have a significantly lower price per carat than a comparable 1.01-carat diamond, because of differences in demand."
Additionally, "Today, because of the relative importance of carat weight among buyers, many diamonds are often intentionally cut poorly to increase carat weight. There is a financial premium for a diamond that weighs the desirable 1.0 carat (200 mg), so often the girdle is made thicker or the depth is increased."
Therefore, there is an incentive to increase carat of the diamond by sacrificing the quality of the cut.
- It seems like lower quality cut diamonds tend to have a higher carat based on the graph, as the median carat for fair quality cuts is approximately 1 carat and is higher than the median for the rest of the cuts. This supports the statement in the Wikipedia article that jewelers prioritize making lower quality cuts to attain higher carat diamonds, particularly diamonds that are 1 milestone, in order to sell it for a higher price.

## Footnotes

[1] Don't mistake my focus on `diamonds` as an endorsement of the diamond
industry. In my opinion aesthetic diamonds are a morally dubious scam.

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-stat00-eda-basics-assignment.Rmd).
